---
title: EDSD - Shapley Decomposition (Exercises) + Solutions
format:
  pdf
---

## Preparation
 
Install the following package from GitHub:

```{r eval=FALSE}
install.packages("remotes")
remotes::install_github("elbersb/shapley")
```

## Exercises – Part 1

1. Proof $\bar{y}_{2}-\bar{y}_{1}=\Delta p+\Delta y$.

**Answer**: Pull out the factor $\frac{1}{2}$ and rewrite as one sum. Expand all terms. The 'mixed'
terms then cancel out and $\bar{y}_{2}-\bar{y}_{1}$ remains.

2. How does the term $\Delta y$ relate to standardization?

**Answer**: $\Delta y$ is the age-standardized difference in the overall means, when
we use $\frac{p_{i1}+p_{i2}}{2}$ as the reference distribution.

3. In R, use the `children` dataset to decompose the difference in overall rates using Kitagawa's method.

```{r}
children <- data.frame(
    age = c("20 to 24", "25 to 29", "30 to 34", "35 to 39", "40 to 44"),
    n1 = c(27, 152, 224, 239, 211),
    rate1 = c(37.037, 19.079, 15.179, 5.021, 6.161),
    n2 = c(363, 208, 96, 59, 48),
    rate2 = c(90.083, 76.923, 56.250, 20.339, 10.417)
)
```

**Answer**: 

```{r}
# create proportions
children$p1 <- children$n1 / sum(children$n1)
children$p2 <- children$n2 / sum(children$n2)

y1 <- with(children, sum(p1 * rate1))
y2 <- with(children, sum(p2 * rate2))
# we want to decompose
y2 - y1

# two terms:
(delta_p <- with(children, sum((rate1 + rate2) / 2 * (p2 - p1))))
(delta_y <- with(children, sum((p1 + p2) / 2 * (rate2 - rate1))))

(delta_p + delta_y == y2 - y1)
```


## Exercises - Part 2

1. What happens if the value function is additive in the factors?
   (e.g., let $N=\{a,b,c\}$, and $v(N)=a+b+c$, $v(\{a,c\})=a+c$, $v({\emptyset})=0$, etc.)

**Answer**: The contribution of factor *a* will simply be $a$, the contribution of factor *b* will simply be $b$, and so on. As $\sum_{S\subseteq N\setminus\{k\}}\tfrac{|S|!(m-1-|S|)!}{m!}=1$ (see exercise below), the Shapley decomposition approach gives the desired answer.

#### Difficult take-home exercises (if you really want to!)

1. Proof that $v(N)-v(\emptyset)=\varphi_{1}+\varphi_{2}+\cdots+\varphi_{m}$. (This is called the efficiency property.)

2. Proof $\sum_{S\subseteq N\setminus\{k\}}\tfrac{|S|!(m-1-|S|)!}{m!}=1$. (This is to see that $\varphi_{i}(N,v)$ is truly a weighted average, with the weights summing to 1.)

## Exercises - Part 3

1. Try the Shapley approach to Kitagawa's decomposition on your own. We defined $v(\emptyset)=\sum_{i}p_{i1}y_{i1}$ and $v(\{\Delta p,\Delta y\})=\sum_{i}p_{i2}y_{i2}$. What happens when you switch these definitions?

2. Rewrite example 3 such that the value function uses a data frame. Try the decomposition on the `austria_chile` data provided below (data from 1981, fertility is per 1,000).

```{r}
austria_chile <- data.frame(
  factor = c("prop_unmarried", "nonm_fertility", "m_fertility"),
  `austria` = c(0.41952, 23.99823, 71.83691),
  `chile` = c(0.475, 50.82674, 115.73732)
)
```

3. Bongaarts (1978) expressed the total fertility rate (TFR) as the product of five factors: the total fecundity rate,
and the indices of proportion married, noncontraception, induced abortion, and lactational infecundability.
The dataset `bongaarts` contains data for South Korea, 1960 and 1970. Use the Shapley value approach to decompose the difference
in the TFR between 1960 and 1970 into the contributions of the five factors.

```{r}
bongaarts <- data.frame(
    factor = c("fecundity", "marriage", "noncontracept", "abortion", "lactation"),
    `1960` = c(16.158, 0.720, 0.970, 0.970, 0.560),
    `1970` = c(16.573, 0.580, 0.760, 0.840, 0.660)
)
```

**Answer**: 

```{r}
library("shapley")

tfr <- function(factors = c()) { 
	v <- bongaarts[["X1960"]] 
	for (i in 1:nrow(bongaarts)) { 
		if (bongaarts[[i, "factor"]] %in% factors) 
			v[i] <- bongaarts[[i, "X1970"]] 
	} 
	prod(v) 
} 

tfr() 
tfr() == prod(bongaarts[["X1960"]])
tfr(c("fecundity", "marriage", "noncontracept", "abortion", "lactation")) 
tfr(c("fecundity", "marriage", "noncontracept", "abortion", "lactation")) == prod(bongaarts[["X1970"]])

shapley(tfr, c("marriage", "noncontracept", "abortion", "lactation", "fecundity"), 
        silent = TRUE) 
```

## Additional exercises

1. Decompose the $R^2$ of a regression model into the contributions of each variable.

2. [Das Gupta (1993)](https://www2.census.gov/library/publications/1993/demographics/p23-186.pdf) provides a
   survey of many decomposition procedures used in demography. Pick a decomposition procedure for which he provides
   data and see whether the Shapley decomposition yields identical results.

3. The dataset `income` below provides information for five individuals on their market income,
   the tax they pay, the transfers they receive, and their final income (final income = market income + tax + transfer).
   The Gini index is an inequality measure: if everyone has the same income, the Gini is 0. If one person has all the income,
   the Gini approaches 1. Taxes and transfers often reduce the Gini, in this case by about 17%. 
   Decompose this difference into the contribution of taxes and transfers.

```{r}
# 100 persons, each has 1 => Gini 0
ineq::Gini(rep(100, 1))
# one person has 100, 100 persons have zero => Gini of 0.99
ineq::Gini(c(100, rep(0, 100)))

income <- data.frame(
	MarketIncome = c(1, 20, 30, 40, 50), 
	Tax1 = c(0, -1, -2, -3, -4), 
	Tax2 = c(-5, -4, -3, -2, -1),
	Transfer = c(9, 7, 5, 3, 1), 
	FinalIncome = c(5, 22, 30, 38, 46)) 
	
ineq::Gini(income[["MarketIncome"]])
ineq::Gini(income[["FinalIncome"]])
```
